---
title: "DESeq2"
author: "Quynh Nhu Nguyen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "Differential Expression Analysis with DESeq2"
author: "Data Analysis Team"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,        # Show code chunks in output
  warning = FALSE,    # Suppress warnings
  message = FALSE,    # Suppress messages
  fig.width = 10,     # Set default figure width
  fig.height = 7      # Set default figure height
)
```

# Required Libraries
```{r libraries}
# Core analysis packages
library(DESeq2)        # For differential expression analysis
library(tidyverse)     # For data manipulation and visualization
library(RColorBrewer)  # For color palettes
library(pheatmap)      # For creating heatmaps
library(EnhancedVolcano) # For volcano plots
library(optparse)      # For command line argument parsing
library(biomaRt)
library(ComplexHeatmap)

```

# Input Data Setup
```{r setup_variables}
# Define input parameters
count_file <- "C:/Users/lucia/OneDrive/Desktop/islam_mouse/human.merged.gene_counts_length_scaled.tsv"
count_col <- 3    # First column containing sample count data
id_col <- 1      # Column containing gene identifiers

# Validate input file existence
if (!file.exists(count_file)) {
  stop("Count file does not exist: ", count_file)
}
```

# Data Import and Processing
```{r data_import}
# Read and process count data
count.table <- read.delim(
  file = count_file, 
  header = TRUE, 
  row.names = NULL,
  check.names = FALSE  # Prevent R from modifying column names
)

# Set gene IDs as row names
rownames(count.table) <- count.table[, id_col]

# Extract only count columns
count.table <- count.table[, count_col:ncol(count.table), drop = FALSE]

# Data quality checks
cat("Dimensions of count table:", dim(count.table), "\n")
cat("Number of NA values:", sum(is.na(count.table)), "\n")

# Show preview of data
head(count.table[, 1:5])
```

# Sample Information Processing
```{r sample_processing}
# Extract sample information from column names
samples.vec <- colnames(count.table)
name_components <- strsplit(samples.vec, "_")

# Create condition information
createConditionData <- function(samples.vec, name_components) {
  n_components <- length(name_components[[1]])
  
  # Create base coldata
  coldata <- data.frame(
    samples.vec, 
    sample = samples.vec, 
    row.names = 1
  )
  
  # Check if sample names can be decomposed
  if (n_components > 1 && all(sapply(name_components, length) == n_components)) {
    # Create groupings based on name components
    groupings <- as.data.frame(
      lapply(1:n_components, function(i) sapply(name_components, "[[", i))
    )
    
    # Keep only informative groupings
    n_distinct <- sapply(groupings, function(grp) length(unique(grp)))
    groupings <- groupings[n_distinct != 1 & n_distinct != length(samples.vec)]
    
    if (ncol(groupings) > 0) {
      names(groupings) <- paste0("CONDITION", 1:ncol(groupings))
      coldata <- cbind(coldata, groupings)
    }
  }
  
  return(coldata)
}

coldata <- createConditionData(samples.vec, name_components)
print(coldata)
```

# DESeq2 Analysis
```{r deseq_analysis}
# Create DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = round(count.table[, samples.vec, drop = FALSE]),
  colData = coldata,
  design = ~ CONDITION1
)

# Perform size factor estimation and differential expression analysis
dds <- estimateSizeFactors(dds)
dds <- DESeq(dds)

# Generate results with different shrinkage estimators
results_list <- list(
  unshrunken = results(dds, contrast = c("CONDITION1", "HIGH", "LOW"), alpha = 0.05),
  apeglm = lfcShrink(dds, coef = "CONDITION1_LOW_vs_HIGH", type = "apeglm"),
  normal = lfcShrink(dds, coef = "CONDITION1_LOW_vs_HIGH", type = "normal"),
  ashr = lfcShrink(dds, coef = "CONDITION1_LOW_vs_HIGH", type = "ashr")
)
```

# Visualization
```{r visualization}
# Create MA plots for each shrinkage method
par(mfrow = c(2, 2))
for (name in names(results_list)) {
  plotMA(results_list[[name]], ylim = c(-2, 2), main = name)
}
```

# Results Analysis
```{r results_analysis}
# Function to analyze differential expression results
analyzeResults <- function(res) {
  # Remove NA values
  res_clean <- res[!is.na(res$padj), ]
  
  # Filter significant genes
  up_genes <- res_clean[res_clean$log2FoldChange > 0 & res_clean$padj < 0.05, ]
  down_genes <- res_clean[res_clean$log2FoldChange < 0 & res_clean$padj < 0.05, ]
  
  # Return summary statistics
  list(
    total = nrow(res),
    clean = nrow(res_clean),
    up = nrow(up_genes),
    down = nrow(down_genes)
  )
}

# Analyze results for normal shrinkage
results_summary <- analyzeResults(results_list$normal)

# Print summary
cat("Analysis Summary:\n",
    "Total genes:", results_summary$total, "\n",
    "Genes after filtering:", results_summary$clean, "\n",
    "Up-regulated genes:", results_summary$up, "\n",
    "Down-regulated genes:", results_summary$down, "\n")
```

# Let’s take a look at the p-value distribution in a histogram
```{r}
results_list$normal %>%
  as.data.frame() %>%
  arrange(padj) %>%
  ggplot(aes(x = pvalue)) +
  geom_histogram(color = "white", bins = 50) +
  labs(
    title = "Histogram of p-values",
    x = "p-value",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r}
# Convert results_list$normal to a tibble and add a column for gene names
results_list_tb <- results_list$normal %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>% 
  as_tibble()

# Connect to the Ensembl database using biomaRt
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Extract the base Ensembl ID (e.g., ENSG00000000419 from ENSG00000000419.8)
results_list_tb <- results_list_tb %>%
  mutate(gene_base = sub("\\..*", "", gene))

# Query gene symbols for each Ensembl ID
genes_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = results_list_tb$gene_base,
  mart = mart
)

# Merge the gene symbols with the original data frame
results_list_tb <- results_list_tb %>%
  left_join(genes_map, by = c("gene_base" = "ensembl_gene_id"))

# View the updated data frame with gene symbols
print(results_list_tb)

```

```{r enhanced_viz}
# Extract significant differentially expressed genes
significant_gene <- results_list_tb %>%
  filter(padj < 0.05 & abs(log2FoldChange) > 0)

# View the significant genes
significant_gene
```
# Enhanced Visualization
```{r enhanced_viz}
# Create Volcano plot
EnhancedVolcano(results_list_tb,
                lab = results_list_tb$hgnc_symbol,
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Differential Expression Analysis',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 4.0,
                col = c('grey30', 'forestgreen', 'royalblue', 'red2'),
                legendLabels = c('NS', 'Log2 FC', 'P value', 'P value & Log2 FC'),
                legendPosition = 'right',
                legendLabSize = 10,
                axisLabSize = 12,
                xlim = c(-5, 5))

EnhancedVolcano(significant_gene,
                lab = significant_gene$hgnc_symbol,
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Differential Expression Analysis in Significant_gene ',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 4.0,
                col = c('grey30', 'forestgreen', 'royalblue', 'red2'),
                legendLabels = c('NS', 'Log2 FC', 'P value', 'P value & Log2 FC'),
                legendPosition = 'right',
                legendLabSize = 10,
                axisLabSize = 12,
                xlim = c(-5, 5))
# Save results to file
write.csv(
  as.data.frame(results_list_tb), 
  file = "differential_expression_results.csv"
)
```

# VST Transformation and ComplexHeatmap
```{r vst_heatmap}
# Load necessary libraries
library(ComplexHeatmap)
library(circlize) # For colorRamp2 function

# Perform VST transformation
vsd <- vst(dds, blind = FALSE)
vst_counts <- assay(vsd)

# Calculate VST statistics
vst_means <- rowMeans(vst_counts)
vst_vars <- rowVars(vst_counts)

# Get results for differential expression
res_df <- as.data.frame(results_list$normal)

# Create gene metrics data frame
gene_metrics <- data.frame(
  mean_exp = vst_means,
  variance = vst_vars,
  padj = res_df$padj,
  log2FC = res_df$log2FoldChange,
  row.names = rownames(vst_counts)
)

# Remove genes with NA values
genes_to_plot <- which(!is.na(gene_metrics$padj))

# Scale the VST counts
mat <- vst_counts[genes_to_plot, ]
mat <- t(scale(t(mat)))  # Scale by gene

# Define gene types
gene_type <- ifelse(
  gene_metrics$padj[genes_to_plot] < 0.05 & 
    abs(gene_metrics$log2FC[genes_to_plot]) > 1,
  "DE", "Non-DE"
)

# Create sample annotation
ha_column = HeatmapAnnotation(
  condition = coldata$CONDITION1,
  mean_expression = anno_boxplot(
    mat,
    outline = FALSE,
    size = unit(2, "mm")
  ),
  col = list(
    condition = c("HIGH" = "#E41A1C", "LOW" = "#377EB8")
  ),
  show_legend = TRUE
)

# Create gene annotation
ha_row = rowAnnotation(
  Expression = anno_boxplot(
    mat,
    outline = FALSE,
    box_width = unit(4, "mm")
  ),
  DE_Status = gene_type,
  log2FC = gene_metrics$log2FC[genes_to_plot],
  col = list(
    DE_Status = c("DE" = "#FF4B4B", "Non-DE" = "#A6A6A6"),
    log2FC = colorRamp2(
      c(-2, 0, 2), 
      c("#377EB8", "white", "#E41A1C")
    )
  ),
  annotation_name_gp = gpar(fontsize = 10),
  width = unit(4, "cm")
)

# Set up colors for heatmap
col_fun = colorRamp2(
  c(min(mat), 0, max(mat)),
  c("#377EB8", "white", "#E41A1C")
)

# Create main heatmap
heatmap_obj <- Heatmap(
  mat,
  name = "VST Z-score",
  column_title = "Samples",
  row_title = "Genes",
  top_annotation = ha_column,
  right_annotation = ha_row,
  
  # Clustering parameters
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  
  # Color parameters
  col = col_fun,
  
  # Cell parameters
  rect_gp = gpar(col = "white", lwd = 0.1),
  
  # Row parameters
  show_row_names = FALSE,
  row_names_gp = gpar(fontsize = 6),
  row_title_gp = gpar(fontsize = 12),
  
  # Column parameters
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 8),
  column_names_rot = 45,
  column_title_gp = gpar(fontsize = 12),
  
  # Split parameters
  row_split = gene_type,
  row_gap = unit(2, "mm"),
  
  # Additional parameters
  show_heatmap_legend = TRUE,
  use_raster = TRUE,
  raster_quality = 2,
  heatmap_legend_param = list(
    title = "VST Z-score",
    legend_height = unit(4, "cm"),
    title_gp = gpar(fontsize = 10, fontface = "bold"),
    labels_gp = gpar(fontsize = 9)
  )
)

# Draw the heatmap
draw(heatmap_obj)


